Investigating soundscapes perception through acoustic scenes simulation.

### 2017-08-23 lafaySimulatedScenePerception.tex
done with JFP comments

\begin{figure}[t]
        \myfloatalign
        \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
\end{figure}

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is. For classes with high level of abstraction that have sub classes, its collection of samples is the union of the collections of the sub classes.

@@

Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is of 1 Volt.

@@

By a volountary design choice, the simulation tool do not allow the user to interact and control a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions. Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane.

@@

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness. For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
\item detect the presence of \emph{outliers} or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as references. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of perceived pleasantness.
\end{enumerate}
\end{itemize}

@@

All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choice concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :
\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, i.e. physically plausible.
\end{itemize}
For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.

@@

44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
Among the 44 participants, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.

@@

All the participants do the experiment on computers with equal hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. All the sound scenes have been re-composed on the basis of the scores obtained in the simulation experiment (Exp. 1.a). The output sound level is the same for all the participants.
All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.

@@

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
For each sound scene, three types of features are considered :
\begin{itemize}
\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in Exp. 1.b, we choose not to normalize the pleasantness scores.
\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
\item \emph{structural}: features related to \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt. The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following tests of significativity:

\begin{itemize}
\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes. The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :
A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.

le test, lui, porte sur la moyenne des Ascene pour les deux catégories
même chose pour les sujets}
\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness. The significativity is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered. The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered. Testing is done at every level of abstraction, and separately for texture and event classes. For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:

@@

\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows us to distinguish between the two types of environment, we investigate the space induced by the semantic features $S$. $S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance. Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or}. The more the composition of the two scenes are similar, the closer they are in this space. The use of Hamming distance allows us to take into account equally the presence and absence of classes. For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named the precision at rank $k$ ($P@k$) . The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$. the $P@k$ is then the average of this ratio for all the items considered as search seeds;

@@

Except for the V-test, the statistical test of significativity are done with a critical threshold of $\alpha=0.05$. For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied. For the $p$-value, if $p\geq0.05$, the value is reported. If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.

@@

First, in order to ensure the coherence of the data, we check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene. Four ni- scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i- scenes. As a consequence, 36 i -scenes and 36 ni -scenes remain for analysis. Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni -scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It appears that the i- scenes were perceived significatively more pleasant ($p < 0.01$) than the ni- scenes.

@@

First, our analyses focuses on the sound level. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$. There is a significative difference in terms of sound level between i- and ni -scenes ($L$: $p<0.01$), with a standard deviation of -7 $dB$. This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and for textures ($L(T)$: $p<0.01$, STD: -6 $dB$).

As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni -scenes tend to be louder, an outcome of a large number of related works. We also notice that this difference of sound level is equally true for events or textures.

It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni- scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significative difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.

Though, the level alone is not sufficient to differentiate the type of environment. In fact, $20\%$ of the i- scenes have a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni- scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni- scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.

When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i- scenes tend to be softer than the ni- scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni- scenes.

@@

\begin{itemize}
\item allows us to differentiate between i- and ni- scenes,
\item characterizes precisely the perceived pleasantness for ni- scenes,
\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i- scenes)
\end{itemize}

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni- scenes). Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involve independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness. Thus, we think that only the level of some of them can influence this percept. In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.

@@

We observe a significant difference between i- and ni- scenes for the classes choice. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.

@@

We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni). Let us now investigate whether some of the sound classes are specific to a given environment. For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

Regarding the sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the participants who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted. Sound markers of ni- scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).

Regarding the textures, 5 markers are identified. For the i- scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}). For the ni- scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).

@@

\begin{table*}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of $V$ value.}
\label{tab:markers}
\end{table*}

@@

Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$. These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes. We also notice that the higher the abstraction level is, the more important the ability to make this distinction is. In other words, the more precise the scene description is, the more clear the discrimination between i- and ni- scenes is.

>>>>

\begin{figure}[t]
        \myfloatalign
        \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
\end{figure}

@@

The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E. Rosch \cite{rosch1978cognition}. The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is. For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.

@@

Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is set arbitrarily to 1 Volt.

@@

By a volountary design choice, the simulation tool does not allow the user to interact and control directly a specific sample. Interaction is done at the track level, a track being a sequence of samples. Several parameters are available to the subject to control the track:

@@

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing. However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potentially induce biases in the analysis conclusions. Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.

@@

Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane, each representing one leaf class.

@@

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}. Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni). Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness. For that, a second group of participants is asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
\item to detect the presence of \emph{outliers} or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as references. We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of perceived pleasantness.
\end{enumerate}
\end{itemize}

@@

All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choices concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints :
\begin{itemize}
\item the listening point of view is the one of a fixed listener;
\item the soundscape must be realistic, i.e. physically plausible. For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
\end{itemize}

These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.

@@

44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.

@@

All the participants do the experiment on computers with equal hardware and software configurations. The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}. The stimuli are the scenes obtained by the simulation experiment of Exp. 1.a. The output sound level is the same for all the participants.
All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.

@@

A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}. In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
For each sound scene, three types of features are considered :
\begin{itemize}
\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in Exp. 1.b, we choose not to normalize the pleasantness scores.
\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
\item \emph{structural}: features related to \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following significance tests:

\begin{itemize}
\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni-scenes is also present in terms of perceived pleasantness, we look at the existence of differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni). The significance is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :
A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.

le test, lui, porte sur la moyenne des Ascene pour les deux catégories
même chose pour les sujets}
\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni-scenes has a significative influence  on the structural features of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness. The significance is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered. The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni-scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered. Testing is done at every level of abstraction, and separately for texture and event classes. For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:

@@

\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows us to distinguish between the two types of environment, we investigate the space induced by the semantic features $S$. $S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance. Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or}. The more the composition of the two scenes are similar, the closer they are in this space. The use of Hamming distance allows us to take into account equally the presence and absence of classes. For measuring the intrinsic ability of the space to split the i- and ni-scene, we use a ranking metric named the precision at rank $k$ ($P@k$) . The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$. the $P@k$ is then the average of this ratio for all the items considered as search seeds;

@@

Except for the V-test, the statistical tests of significance are done with a critical threshold of $\alpha=0.05$. For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied. For the $p$-value, if $p\geq0.05$, the value is reported. If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.

@@

First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene. Four ni-scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i-scenes. As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis. Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni-scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It appears that the i-scenes were perceived significatively more pleasant ($p < 0.01$) than the ni-scenes.

@@

First, our analyses focuses on the sound level. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$. There is a significative difference in terms of sound level between i- and ni-scenes ($L$: $p<0.01$). This difference is siginificant for events ($L(E)$: $p<0.01$) and for textures ($L(T)$: $p<0.01$).
%, with a standard deviation of -7 $dB$
%, STD: -7 $dB$
%, STD: -6 $dB$

As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni-scenes tend to be louder. This results is also an outcome of a large number of related works. We also notice that this difference of sound level is both significant for events or textures.

It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significative difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.

Though, the level alone is not sufficient to differentiate the type of environment. In fact, $20\%$ of the i-scenes have a sound level higher than the nominal level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni-scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.

When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni-scenes.

@@

\begin{itemize}
\item allows us to differentiate between i- and ni-scenes,
\item characterizes precisely the perceived pleasantness for ni-scenes,
\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i-scenes)
\end{itemize}

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes). Thus, the sound events keep a relative ability to predict the pleasantness of the ni-scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involves independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the i-scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness. Thus, we think that only the level of some of them can influence this percept. In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.

@@

We observe a significant difference between i- and ni-scenes for the classes choice. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i-scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.

@@

We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni). Let us now investigate whether some of the sound classes are specific to a given environment. For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level. Results are presented in Table~\ref{tab:markers}. \jfp{le V-test est un test statistique : on peut mettre la p-value dans le tableau}

Regarding the sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the participants who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted. Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).

Regarding the textures, 5 markers are identified. For the i-scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}). For the ni-scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).

@@

\begin{table*}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value.\jfp{a partir de quelle valeur de V-test le marker est présent dans ce tableau ?

mettre la p-value , non ? }}
\label{tab:markers}
\end{table*}

@@

Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$. These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes. We also notice that the higher the abstraction level is, the more important the ability to make this distinction is. In other words, the more precise the scene description is, the more clear the discrimination between i- and ni-scenes is.
